A thermodynamic view of dusty protoplanetary disks

### 2017-06-30 paper.tex
more mods to introduction.  need to proofread before MK judges...

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{goldreich1973, chiang08,barranco09,lee10}. These instabilities 
can trigger turbulence, and (in different regimes) promote or inhibit 
planetesimal formation. 
More recently, new 
dusty instabilities have appeared in numerical
simulations \citep{loren15,loren16,lamb16}, but their physical underpinnings are not yet well
understood.

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagragian dust particles
\citep{johansen2006, nelson10,bai10,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

Another approach is to model the dust population as a continuous, pressureless
fluid 
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of existing numerical methods/codes for
simulating pure gas dynamics.

@@

It is possible to re-formulate the two-fluid equations into that for a
single fluid with total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the relative dust-gas velocity
$\Delta\bm{v}\equiv \bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$. The latter also contributes to 
additional source terms in the momentum and energy equations.

The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for two-fluid models and 
following Lagragian particles; but is straight forward with the 
one-fluid formulation \citep{price15}.  The single fluid framework 
also has the important advantage that it is 
amenable to standard analytical methods \citep{youdin05a,jacquet11}.  
This is the necessary approach to gain physical insight.

In this work, we are motivated to develop a general analytical
framework to study the stability of dusty protoplanetary 
disks by drawing on the physical and mathematical parallel between
dusty gas and standard hydrodynamics. This is possible in certain limits of astrophysical
interest, allowing us to interpret dusty phenomena through standard
hydrodynamics.

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set the dust-gas relative drift
to $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a, 
  jacquet11,laibe14}. This approximation reflects the physical result
that dust particles tend to drift towards pressure maximum.

 We also consider irradiated protoplanetary disks 
 in which the temperature is set externally
\citep{chiang97,stam08}. This allows us to adopt a locally isothermal
equation of state, $P = c_s^2\rhog$, where $c_s$ is a prescribed
sound-speed profile. This physically represents short cooling times
\citep{lin15}.

Under these two approximations, we show that the equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics (Fig. \ref{concept_chart}). This is because tightly-coupled dust particles are almost
entrained in the gas, which is analogous to the advection
of entropy following an adiabatic fluid. We show that the   
effective entropy of isothermal dusty gas scales with its gas
fraction. Finite dust-gas drag act as effective heating/cooling: the relative
drift between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into standard  
hydrodynamics, we can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations.%, regardless of how  thin the dust
%layer is. 
{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. We first present, without
derivation, the correspondence between dusty gas and     
ideal hydrodynamics in \S\ref{key_idea}. In \S\ref{grow_osc} 
we discuss a key idea in fluid dynamics, relevant to our 
discussion on  SI and VSI: instabilities associated with
pressure-density phase lags.   
%This means that dusty-gas phenomena, such 
%as SI and VSI, have thermodynamical interpretations.

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we list the linearized equations and show that 
they reproduce previous results for SGI and SI. In \S\ref{results} we
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

>>>>

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{goldreich73, chiang08,barranco09,lee10}. These instabilities 
can trigger turbulence, and (in different regimes) promote or inhibit 
planetesimal formation. 
More recently, new 
dusty instabilities have appeared in numerical
simulations \citep{loren15,loren16,lamb16}, but their physical underpinnings are not yet well
understood.

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagragian dust particles
\citep{johansen06, nelson10,bai10,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

Another approach is to model the dust population as a continuous, pressureless
fluid 
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of existing numerical methods/codes for
simulating pure gas dynamics. A practical difficulty with this approach is the need to numerical stabilize the pressureless dust fluid with (e.g.) artificial diffusion.

@@

%AY: reworded b/c (a) I didn't think it was clear enough that the one fluid simplification wasn't exact and (b) that this tight coiupling limit needed to be emphasized more.

The two-fluid equations for dust and gas can be reformulated into 
equivalent dynamical equations for the center of mass velocity $\bm{v}$
and the relative dust-gas velocity  $\Delta\bm{v}\equiv \bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$ 
\citep{youdin05a}. In this reformulation, continuity equations for dust and gas can be replaced 
by the mass conservation of total density $\rho$ plus the evolution of the dust-to-gas ratio 
(or dust fraction, \citealp{laibe14}).

The advantage of this reformulation arises in the tight coupling limit of  $\tstop \ll 1/\Omega$, where the orbital frequency $\Omega$ sets the characteristic timescale in many disk dynamics problems.
Models with two fluids or with Lagrangian solids are numerically difficult to evaluate in this regime, due to stiff drag forces.  
In the reformulated equations, however, the center of mass motion does not experience drag forces and thus is not stiff.  Moreover the relative motion satisfies the terminal-velocity approximation and can be eliminated from the equations.  This single fluid framework for a well-coupled dust-gas mixture facilitates numerical calculations \citep{price15}. and also simplifies analytic calculations \citep{youdin05a,jacquet11}, which is useful for gaining insight.

%It is possible to re-formulate the two-fluid equations into that for a
%single fluid with total density $\rho$ and center-of-mass
%velocity $\bm{v}$ of the dusty gas mixture
%\citep{laibe14}. The standard hydrodynamic equations for
%$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
%the dust-to-gas ratio $\rhod/\rhog$ and the relative dust-gas velocity
%$\Delta\bm{v}\equiv \bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$. The latter also contributes to 
%additional source terms in the momentum and energy equations.  
%
%The one-fluid approach is appropriate for 
%simulating gas tightly coupled to dust particles with small $\tstop$.  
%This can be numerically challenging for two-fluid models and 
%following Lagragian particles; but is straight forward with the 
%one-fluid formulation \citep{price15}.  The single fluid framework 
%also has the important advantage that it is 
%amenable to standard analytical methods \citep{youdin05a,jacquet11}.  
%This is the necessary approach to gain physical insight.

%AY: Attempted to clarify a bit by separating more the analogy and it's applications.

In this work, we develop a parallel between this single-fluid description of dust-gas mixtures and standard hydrodynamics.
Specifically we recast the evolution of the dust-to-gas ratio as an energy equation for the thermal content of a fluid.
This parallel is most precise when the gas in the dust/gas mixture obeys a locally isothermal equation of state for the gas, $P = c_s^2\rhog$, where $c_s$ is a prescribed sound-speed profile.  Thus we make this approximation which holds when cooling times are short \citep{lin15}, as with protoplanetary disks 
 whose temperature is set by external irradiation
\citep{chiang97,stam08}.

%are motivated to develop a general analytical
%framework to study the stability of dusty protoplanetary 
%disks by drawing on the physical and mathematical parallel between
%dusty gas and standard hydrodynamics. This is possible in certain limits of astrophysical
%interest, allowing us to interpret dusty phenomena through standard
%hydrodynamics.

%AY: a bit detailed for an introduction, plus the idea of terminal velocity needed to be introduced earlier as the basis of the slngle fluid equations, I thought.
%We consider tightly coupled dust particles with $\tstop\Omega\ll
%1$, where $\Omega$ is the disk orbital frequency. We may then use the
%`terminal velocity approximation' to set the dust-gas relative drift
%to $\Delta \bm{v} = \tstop\nabla
%P/\rhog$, where $P$ is the gas pressure \citep{youdin05a, 
%  jacquet11,laibe14}. This approximation reflects the physical result
%that dust particles tend to drift towards pressure maximum.

% We also consider irradiated protoplanetary disks 
% in which the temperature is set externally
%\citep{chiang97,stam08}. This allows us to adopt a locally isothermal
%equation of state, $P = c_s^2\rhog$, where $c_s$ is a prescribed
%sound-speed profile. This physically represents short cooling times
%\citep{lin15}.

Under these two approximations of strong drag and isothermal gas, we show that the equations of 
 dusty gas dynamics are equivalent to that of nearly adiabatic (i.e. slowly cooling) gas
dynamics without dust (Fig. \ref{concept_chart}).  The basis of the analogy is that the entrainment 
of tightly-coupled dust particles in gas flows is similar to the advection
of entropy in an adiabatic fluid. %We show that the   
%effective entropy of isothermal dusty gas scales with its gas
%fraction.  %AY: too technical for introduction.
Finite dust-gas drag act as effective heating/cooling: the relative
drift between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.   Needless to say the effective heating term must have a 
specific form, which we derive, for the equivalency to hold.

@@

The purpose of our physical and mathematical parallel between
dusty gas and standard hydrodynamics is to gain a deeper understanding of the dynamics
and stability of dusty disks.  
We can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations.%, regardless of how  thin the dust
%layer is. 
{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. 
%AY: or not? %We first present, without
%derivation, the correspondence between dusty gas and     
%ideal hydrodynamics in \S\ref{key_idea}. 
In \S\ref{grow_osc} 
we describe a key insight of our study, that pressure (i.e. ``PdV") work explains the basis of several instabilities, including SI and VSI, in terms of
pressure-density phase lags.   
%This means that dusty-gas phenomena, such 
%as SI and VSI, have thermodynamical interpretations.  
We begin our formal discussion in \S\ref{setup} by transforming the
basic equations describing tightly-coupled dusty gas into  hydrodynamic equations with a special cooling function. 
We derive equilibrium solutions in \S\ref{eqm}, which we use in our explicit stability calculations. In \S\ref{limits} we analyze the 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we show that our linearized equations 
reproduce the SGI and SI in the tight coupling regime. In \S\ref{results} we
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.
